The Unicode Cookbook \linebreak for Linguists

### 2018-06-08 book/chapters/ipa_background.tex
paperhivers comments through chapter 3

In this chapter we present a brief history of the IPA 
(Section~\ref{IPAhistory}), which itself dates 
back to the late 19th century, not long after the creation of the first 
typewriter with a QWERTY 
keyboard. An understanding of the IPA and its premises and principles 
(Section~\ref{IPApremises-principles}) leads to a better 
appreciation of the challenges that the International Phonetic Association 
faced when digitally encoding the IPA's set of symbols and 
diacritics (Section~\ref{EncodingIPA}). Occurring a bit over one hundred 
years after the IPA's inception, the encoding challenge was no small feat 
(Section~\ref{need-for-multilingual-environment}); many 
linguists have encountered the pitfalls when the twain shall meet 
(Chapter~\ref{ipa-meets-unicode}).

@@

Originally, the IPA was a list of symbols with pronunciation examples
using words in different languages. In 1900 the symbols were first organized into
a chart and were given phonetic feature labels (e.g.\ for manner of
articulation among others ``plosives'', ``nasales'', ``fricatives'', for place of
articulation among others ``bronchiales'', ``laryngales'', ``labiales'' and for vowels
e.g.\ ``fermées'', ``mi-fermées'', ``mi-ouvertes'', ``ouvertes''). Throughout the last
century, the structure of the chart has changed with increases in phonetic
knowledge. Thus, similar to notational systems in other scientific disciplines,
the IPA reflects facts and theories of phonetic knowledge that have developed
over time. It is natural then that the IPA is modified occasionally to
accommodate scientific innovations and discoveries. In fact, updates are part of the
Association's mandate. These changes are captured in revisions to the IPA chart.\footnote{For a detailed history see:
\url{https://en.wikipedia.org/wiki/History\_of\_the\_International\_Phonetic_Alphabet}}

Over the years there have been several revisions, but mostly minor. Articulation 
labels -- what are often called `features' even though the IPA
deliberately avoids this term -- have changed (e.g.\ terms like ``lips'', ``throat''
or ``rolled'' are not used anymore). Phonetic symbol values have changed (e.g.\
voiceless is no longer marked by <h>). Symbols have been dropped (e.g.\ the
caret diacritic denoting `long and narrow' is no longer used). And many symbols
have been added to reflect contrastive sounds found in the world's very diverse
phonological systems. The use of the IPA is guided by principles outlined in 
the \textsc{handbook of the international phonetic association} \citep{IPA1999}, 
henceforth simply called \textit{Handbook}.

@@

	\item it is intended to be a set of symbols for representing all possible
       sounds in the world's (spoken) languages;
	\item its chart has a linguistic basis (specifically a phonological bias)
       rather than just being a general phonetic notation scheme;
	\item its symbols can be used to represent distinctive feature
       combinations;\footnote{Although the chart uses traditional manner and
       place of articulation labels, the symbols can be used as a representation
       of any defined bundle of features, binary or otherwise, to define
       phonetic dimensions.}
	\item its chart provides a summary of linguists' agreed-upon phonetic 
	knowledge.

@@

Once spoken language data are segmented, the IPA provides symbols to
unambiguously represent phonetic details. However, since phonetic detail could
potentially include anything, e.g.\ something like ``deep voice'', the IPA
restricts phonetic detail to linguistically relevant aspects of speech.
Phonological consideration thus inextricably play a roll in transcription. In
other words, phonetic observations beyond quantitative acoustic analysis are
always made in terms of some phonological framework.

Today, the IPA chart reflects a linguistic theory grounded in principles of
phonological contrast and in knowledge about the attested linguistic variation.
This fact is stated explicitly in several places, including in the
\textit{Report on the 1989 Kiel convention} published in the \textit{Journal of
the International Phonetic Association} \citep[67-68]{International1989report}:

@@

\noindent From its earliest days the Association aimed to provide ``a separate sign for
each distinctive sound; that is, for each sound which, being used instead of
another, in the same language, can change the meaning of a word''
\citep[27]{IPA1999}. Distinctive sounds became later known as \textsc{phonemes}
and the IPA has developed historically into a notational devise with a strictly
segmental phonemic view. A phoneme is an abstract theoretical notion derived
from an acoustic signal as produced by speakers in the real world. Therefore the
IPA contains a number of theoretical assumptions about speech and how to
transcribe speech in written form.

@@

Essentially transcription has two parts: a text containing symbols and a set 
of conventions for interpreting those symbols (and their combinations). 
The symbols of the IPA distinguish between letter-like symbols and
diacritics (symbol modifiers). The use of the letter-like symbols to represent 
a language's sounds is guided by the principle of contrast; where two words 
are distinguishable by phonemic contrast, those contrasts should be transcribed 
with different letter symbols (and not just diacritics). Allophonic distinction 
falls under the rubric of diacritically-distinguished symbols, e.g.\ the 
difference in English between an aspirated /p/ and /t/ in [spʰot̚] and 
an unreleased /p/ in [stʰop̚].

>>>>

In this chapter we present a brief history of the IPA 
(Section~\ref{IPAhistory}), which dates 
back to the late 19th century, not long after the creation of the first 
typewriter with a QWERTY 
keyboard. An understanding of the IPA and its premises and principles 
(Section~\ref{IPApremises-principles}) leads to a better 
appreciation of the challenges that the International Phonetic Association 
faced when digitally encoding the IPA's set of symbols and 
diacritics (Section~\ref{EncodingIPA}). Occurring a little over a hundred years after 
the inception of the IPA, its encoding was a major challenge 
(Section~\ref{need-for-multilingual-environment}); many 
linguists have encountered the pitfalls when the two are used together 
(Chapter~\ref{ipa-meets-unicode}).

@@

Originally, the IPA was a list of symbols with pronunciation examples
using words in different languages. In 1900 the symbols were first organized into
a chart and were given phonetic feature labels, e.g.\ for manner of
articulation among others \textit{plosives}, \textit{nasales}, \textit{fricatives}, for place of
articulation among others \textit{bronchiales}, \textit{laryngales}, \textit{labiales} and for vowels
e.g.\ \textit{fermées}, \textit{mi-fermées}, \textit{mi-ouvertes}, \textit{ouvertes}. Throughout the last
century, the structure of the chart has changed with increases in phonetic
knowledge. Thus, similar to notational systems in other scientific disciplines,
the IPA reflects facts and theories of phonetic knowledge that have developed
over time. It is natural then that the IPA is modified occasionally to
accommodate scientific innovations and discoveries. In fact, updates are part of the
Association's mandate. These changes are captured in revisions to the IPA chart.\footnote{For a detailed history, 
we refer the reader to:
\url{https://en.wikipedia.org/wiki/History\_of\_the\_International\_Phonetic_Alphabet}.}

Over the years there have been several revisions, but mostly minor ones. Articulation 
labels -- what are often called \textit{features} even though the IPA
deliberately avoids this term -- have changed, e.g.\ terms like \textit{lips}, \textit{throat}
or \textit{rolled} are no longer used. Phonetic symbol values have changed, e.g.\
voiceless is no longer marked by <h>. Symbols have been dropped, e.g.\ the
caret diacritic denoting `long and narrow' is no longer used. And many symbols
have been added to reflect contrastive sounds found in the world's very diverse
phonological systems. The use of the IPA is guided by principles outlined in 
the \textit{Handbook of the International Phonetic Association} \citep{IPA1999}, 
henceforth simply called \textit{Handbook}.

@@

	\item it is intended to be a set of symbols for representing all possible
       sounds in the world's (spoken) languages;
	\item its chart has a linguistic basis (and specifically a phonological bias)
       rather than just being a general phonetic notation scheme;
	\item its symbols can be used to represent distinctive feature
       combinations;\footnote{Although the chart uses traditional manner and
       place of articulation labels, the symbols can be used as a representation
       of any defined bundle of features, binary or otherwise, to define
       phonetic dimensions.}
	\item its chart provides a summary of linguists' agreed-upon phonetic 
	knowledge.

@@

Once spoken language data are segmented, the IPA provides symbols to
unambiguously represent phonetic details. However, since phonetic detail could
potentially include anything, e.g.\ something like ``deep voice'', the IPA
restricts phonetic detail to linguistically relevant aspects of speech.
Phonological considerations thus inextricably play a roll in transcription. In
other words, phonetic observations beyond quantitative acoustic analysis are
always made in terms of some phonological framework.

Today, the IPA chart reflects a linguistic theory grounded in principles of
phonological contrast and in knowledge about the attested linguistic variation.
This fact is stated explicitly in several places, including in the
\textit{Report on the 1989 Kiel convention} published in the \textit{Journal of
the International Phonetic Association} \citep[67--68]{International1989report}:

@@

\noindent From its earliest days the Association aimed to provide ``a separate sign for
each distinctive sound; that is, for each sound which, being used instead of
another, in the same language, can change the meaning of a word''
\citep[27]{IPA1999}. Distinctive sounds became later known as \textsc{phonemes}
and the IPA has developed historically into a notational device with a strictly
segmental phonemic view. A phoneme is an abstract theoretical notion derived
from an acoustic signal as produced by speakers in the real world. Therefore the
IPA contains a number of theoretical assumptions about speech and how to
transcribe speech in written form.

@@

Essentially, transcription has two parts: a text containing symbols and a set 
of conventions for interpreting those symbols (and their combinations). 
The symbols of the IPA distinguish between letter-like symbols and
diacritics (symbol modifiers). The use of the letter-like symbols to represent 
a language's sounds is guided by the principle of contrast; where two words 
are distinguishable by phonemic contrast, those contrasts should be transcribed 
with different letter symbols (and not just diacritics). Allophonic distinction 
falls under the rubric of diacritically-distinguished symbols, e.g.\ the 
difference in English between an aspirated /p/ and /t/ in [spʰot̚] and 
an unreleased /p/ in [stʰop̚].
